\clearpage

\begin{thebibliography}{99}

\bibitem[HarperHonsellPlotkin93]{HarperHonsellPlotkin93}
Robert Harper, Furio Honsell, Gordon Plotkin.
\textit{A framework for defining logics}.
In \textit{Journal of the ACM (JACM), v.40 n.1}, pages 143-184, 1993.

\bibitem[Pfenning91]{Pfenning91}
Frank Pfenning.
\textit{Logic programming in the LF logical framework}.
Cambridge University Press, 1991.

\bibitem[PfenningSch\"{u}rmann99]{PfenningSchurmann99}
Frank Pfenning, Carsten Sch\"{u}rmann.
\textit{System Description: Twelf - A Meta-Logical Framework for Deductive Systems}.
In \textit{Proceedings of the 16th International Conference on Automated Deduction: Automated Deduction}, pages 202-206, 1999.

\bibitem[Pfenning01]{Pfenning01}
Frank Pfenning.
\textit{Computation and Deduction}.
Unpublished course notes, 2001.
(To be published by Cambridge University Press, March 2016.)

\bibitem[Twelf16]{Twelf16}
Twelf home page. Available at \texttt{http://www.twelf.org/}, 2016.

\bibitem[Hannan92]{Hannan92}
John Hannan and Frank Pfenning.
\textit{Compiler verification in LF}.
In \textit{Seventh Annual IEEE Symposium on Logic in Computer Science}, pages 407–-418, 1992.

\bibitem[Simmons07]{Simmons07}
Robert J. Simmons.
\textit{Concrete Representation}.
Available at \texttt{http://twelf.org/wiki/Concrete\_representation}, 2016 (last updated 2007).

\bibitem[Crary08]{Crary08}
Karl Crary.
\textit{Explicit Contexts in LF}.
In \textit{Workshop on Logical Frameworks and Meta-Languages: Theory and Practice}, 2008 (revised 2009).

\bibitem[CompCert16]{CompCert16}
CompCert home page. \\
Available at \texttt{http://compcert.inria.fr/}, 2016.

\bibitem[Chlipala10]{Chlipala10}
Adam Chlipala.
\textit{A Verified Compiler for an Impure Functional Language},
In \textit{Proceedings of the 37th ACM SIGPLAN-SIGACT Symposium on Principles of Programming Languages}, pages 93-106, 2010.

\bibitem[Yang11]{Yang11}
Xuejun Yang, Yang Chen, Eric Eide, John Regehr.
\textit{Finding and understanding bugs in C compilers}
In \textit{Proceedings of the 32nd ACM SIGPLAN conference on Programming language design and implementation}, pages 283-294, 2011.

\bibitem[EideRegehr08]{EideRegehr08}
Eric Eide, John Regehr.
\textit{Volatiles Are Miscompiled, and What to Do about It}.
In \textit{Proceedings of the 8th ACM international conference on Embedded software}, pages 255-264, 2008.

\bibitem[AppelJim88]{AppelJim88}
Andrew W. Appel, Trevor T. Y. Jim.
\textit{Optimizing closure environment representations}.
Technical Report 168, Dept. of Computer Science, Princeton University, 1988.

\bibitem[Myreen10]{Myreen10}
Magnus O. Myreen.
\textit{Reusable verification of a copying collector}.
In \textit{Proceedings of the Third international conference on Verified software: theories, tools, experiments}, pages 142-156, 2010.

\end{thebibliography}
